---
title: "Project1_SM"
author: "Sofia Mariano"
date: "30 Sept. 2026"
format: html
editor: visual
---
Prepare session to perform analysis:
```{r}
library(httr)
library(jsonlite)
library(tidyverse)
library(dplyr)
library(lubridate)
library(stringr)
```

# Part 1: Data Processing
## Parsing Variable Values
Create list for 2021
```{r}
# Read in raw 2021 data
data_2021 <- httr::GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2021 <- jsonlite::fromJSON(rawToChar(data_2021$content))
variables_2021 <- parsed_2021$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION")]
```

Create list for 2022
```{r}
# Read in raw 2022 data
data_2022 <- httr::GET("https://api.census.gov/data/2022/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2022 <- jsonlite::fromJSON(rawToChar(data_2022$content))
variables_2022 <- parsed_2022$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION")]
```

Create list for 2023
```{r}
# Read in raw 2023 data
data_2023 <- httr::GET("https://api.census.gov/data/2023/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2023 <- jsonlite::fromJSON(rawToChar(data_2023$content))
variables_2023 <- parsed_2023$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION")]
```

Create list for 2024
```{r}
# Read in raw 2024 data
data_2024 <- httr::GET("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2024 <- jsonlite::fromJSON(rawToChar(data_2024$content))
variables_2024 <- parsed_2024$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION")]
```

Combine individual year variable lists into a single RDS file
```{r}
# Save all individual year variable lists into a single list
combined_variables <- c(variables_2021, variables_2022, variables_2023, variables_2024)

# Save combined list as a single RDS file
saveRDS(combined_variables, file = "data/rawdata_variables.RDS")
```

## Parsing API Data
Using usual coding practice, query API and return a tibble.  Data from 2021 used for this example.
```{r}
# Read-in URL
usual_2021 <- httr::GET("https://api.census.gov/data/2021/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=b195a5a7b89af4ff112dcbfe5e6a9eff2102afad")

# Parse data
parsed_usual_2021 <- jsonlite::fromJSON(rawToChar(usual_2021$content))

# Return as a tibble
tibble_2021 <- dplyr::as_tibble(parsed_usual_2021)

# Print tibble
tibble_2021

# Column names aren't what they should be in the first tibble.  Fix tibble output so the first row now acts as the column names.
colnames(tibble_2021) <- tibble_2021[1,]
tibble_2021 <- tibble_2021[-1,]

# Print tibble
tibble_2021

# The output variables aren't the appropriate type.  Fix variables so that their type matches the predicate type listed in the variables definition link.
tibble_2021 <- tibble_2021 |>
    dplyr::mutate(SEX <- as.factor(SEX),
                PWGTP <- as.numeric(PWGTP),
                MAR <- as.factor(MAR),
                SCHL <- as.factor(SCHL)
                ) |>
    dplyr::select(`SEX <- as.factor(SEX)`, `PWGTP <- as.numeric(PWGTP)`, `MAR <- as.factor(MAR)`, `SCHL <- as.factor(SCHL)`) |>
    dplyr::rename("SEX" = "SEX <- as.factor(SEX)", "PWGTP" = "PWGTP <- as.numeric(PWGTP)", "MAR" = "MAR <- as.factor(MAR)", "SCHL" = "SCHL <- as.factor(SCHL)")

# Print tibble to check output
tibble_2021
```

Create a single function that would combine all three processes above that would directly return a given URL as a tibble
```{r}
search_census <- function(URL){
  result <- httr::GET(URL)
  parsed <- jsonlite::fromJSON(rawToChar(result$content))
  tibble_census_test <- dplyr::as_tibble(parsed)
  colnames(tibble_census_test) <- tibble_census_test[1,]
  tibble_census_test <- tibble_census_test[-1,]
  tibble_census_test <- tibble_census_test |>
    dplyr::mutate(SEX <- as.factor(SEX),
                PWGTP <- as.numeric(PWGTP),
                MAR <- as.factor(MAR),
                SCHL <- as.factor(SCHL)
                ) |>
    dplyr::select(`SEX <- as.factor(SEX)`, `PWGTP <- as.numeric(PWGTP)`, `MAR <- as.factor(MAR)`, `SCHL <- as.factor(SCHL)`) |>
    dplyr::rename("SEX" = "SEX <- as.factor(SEX)", "PWGTP" = "PWGTP <- as.numeric(PWGTP)", "MAR" = "MAR <- as.factor(MAR)", "SCHL" = "SCHL <- as.factor(SCHL)")
  return(tibble_census_test)
}
```

Test function using example data from 2023.
```{r}
# Set URL variable with example data 
URL_2023 <- "https://api.census.gov/data/2023/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=20&key=b195a5a7b89af4ff112dcbfe5e6a9eff2102afad"

# Test function
search_census(URL_2023)
```

Write a function that would query an API while also parsing the values.
```{r}
# Call raw data variable file in order to reference in function
variable_list <- readRDS("data/rawdata_variables.RDS")

# Create function
query_census <- function(year = "2024", numeric1 = "PWGTP", numeric2 = "AGEP", categorical1 = "SEX", geography = "STATE", geography_value = "24",key = "b195a5a7b89af4ff112dcbfe5e6a9eff2102afad"){
  # year variable formatting
  if (year %in% c("2021","2022","2023","2024")) {
    print(year)
  } else {
    "Error: Incorrect Year"
  }
  # numeric variable formatting
  if (numeric1 %in% c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", 
"JWMNP","PWGTP")) {
    print(numeric1)
  }
  if (numeric2 %in% c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", 
"JWMNP","PWGTP")) {
    print(numeric2)
  } 
  for (item in variable_list$JWAP$item){
    if (is.na(variable_list$JWAP$item)){
      "NA"
    } else if (!is.na(variable_list$JWAP$item)){
      extracted_range <- stringr::str_match(variable_list$JWAP$item, ':')[,2] #Split into individual values following item number
      #select out N/A values for 0
      if (extracted_range != "to"){
        print(extracted_range)
      } else {calculate_midpoint <- function(extracted_range)
      # Split string into start and end time
      time_parts <- stringr::str_split(extracted_range, "to")[]
    # Parse individual times
    start_time <- readr::parse_time(time_parts)
    end_time <- readr::parse_time(time_parts)
    # Convert seconds since midnight
    start_seconds <- as.numeric(start_time) %% (24*60*60)
    end_seconds <- as.numeric(end_time) %% (24*60*60)
    # Calculate midpoint in seconds
    midpoint_seconds <- (start_seconds + end_seconds) / 2
    # Convert midpoint to time object
    midpoint_time <- as.POSIXct(midpoint_seconds, tz = "UTC")
    # Format to display time only
    midpoint_format <- format(midpoint_time, format = "%I:%M %p")
    return(midpoint_format)
      }
    }
  }
  for (item in variable_list$JWDP$item){
    if (is.na(variable_list$JWDP$item)){
      "NA"
    } else if (!is.na(variable_list$JWDP$item)){
      extracted_range <- stringr::str_match(variable_list$JWDP$item, ':')[,2] #Split into individual values following item number
      #select out N/A values for 0
      if (extracted_range != "to"){
        print(extracted_range)
      } else {calculate_midpoint <- function(extracted_range)
      # Split string into start and end time
      time_parts <- stringr::str_split(extracted_range, "to")[]
    # Parse individual times
    start_time <- readr::parse_time(time_parts)
    end_time <- readr::parse_time(time_parts)
    # Convert seconds since midnight
    start_seconds <- as.numeric(start_time) %% (24*60*60)
    end_seconds <- as.numeric(end_time) %% (24*60*60)
    # Calculate midpoint in seconds
    midpoint_seconds <- (start_seconds + end_seconds) / 2
    # Convert midpoint to time object
    midpoint_time <- as.POSIXct(midpoint_seconds, tz = "UTC")
    # Format to display time only
    midpoint_format <- format(midpoint_time, format = "%I:%M %p")
    return(midpoint_format)
      }
    }
  }
  if ((numeric1 != "PWGTP") | (numeric2 != "PWGTP")){
    "Error: Missing Key Numeric Variable PWGTP"
  }
  #categorical variable formatting
  if (categorical1 %in% c("FER", "HHL", "SCH", "SCHL", "SEX")){
    print(categorical1)
  } else {
    "Error: Incorrect categorical variable specified"
  }
  # geographic variable formatting
  if (geography %in% c("REGION", "DIVISION", "ST", "STATE")) {
    print(geography)
  } else {
    "Error: Incorrect Geography Level"
  }
  if (geography_value == "all") {
    print(geography_value)
  } else if (geography == "REGION" & geography_value %in% variable_list$REGION) {
    print(geography_value)
  } else if (geography == "DIVISION" & geography_value %in% variable_list$DIVISION) {
    print(geography_value)
  } else if (geography == "ST" & geography_value %in% variable_list$ST) {
    print(geography_value)
  } else if (geography == "STATE" & geography_value %in% variable_list$STATE){
    print(geography_value)
  } else {
    "Error: Incorrect Geography Subset Specified"
  }

# coding for URL begins after formatting completed
  result <- httr::GET(paste0("https://api.census.gov/data/",
                             year,
                             "/acs/acs1/pums?get=",
                             numeric1,
                             ",",
                             numeric2,
                             ",",
                             categorical1,
                             "&",
                             geography,
                             "=",
                             geography_value,
                             "&key=",
                             key
  ))
  parsed <- jsonlite::fromJSON(rawToChar(result$content))
  tibble_census <- dplyr::as_tibble(parsed)
  colnames(tibble_census) <- tibble_census[1,]
  tibble_census <- tibble_census[-1,]
  tibble_census <- tibble_census |>
    dplyr::mutate(PWGTP <- as.numeric(PWGTP),
                  AGEP <- as.numeric(AGEP),
                  GASP <- as.numeric(GASP),
                  GRPIP <- as.numeric(GRPIP),
                  JWMNP <- as.numeric(JWMNP),
                  FER <- factor(FER, 0:2, c( "N/A (less than 15 years/greater than 50 years/ male)", "No", "Yes")
                     ),
                  SEX <- factor(SEX,1:2, c("Male","Female")
                     ),
                  SCH <- factor(SCH,0:3, c("N/A (less than 3 years old)", "No, has not attended in the last 3 months", "Yes, public school or public college", "Yes, private school or college or home school")
                     ),
                  SCHL <- factor(SCHL,0:24, c("N/A (less than 3 years old)", "No schooling completed", "Nursery school, preschool", "Kindergarten", "Grade 1", "Grade 2", "Grade 3", "Grade 4", "Grade 5", "Grade 6", "Grade 7", "Grade 8", "Grade 9", "Grade 10", "Grade 11", "12th grade - no diploma", "Regular high school diploma", "GED or alternative credential", "Some college, but less than 1 year", "1 or more years of college credit, no degree", "Associate's degree", "Bachelor's degree", "Master's degree", "Professional degree beyond a bachelor's degree", "Doctorate degree")
                      )
                ) |>
    dplyr::rename("PWGTP" = "PWGTP <- as.numeric(PWGTP)", "AGEP" = "AGEP <- as.numeric(AGEP)", "GASP" = "GASP <- as.numeric(GASP)", "GRPIP" = "GRPIP <- as.numeric(GRPIP)", "JWMNP" = "JWMNP <- as.numeric(JWMNP)")
  class(parsed) <- c("census", class(parsed))
  return(tibble_census)
}
```

Testing the function in various conditions
```{r}
# Testing defaults
query_census()

# Testing acceptable years
query_census("2023")

# Testing unacceptable years
query_census("2026")

# Testing acceptable numeric variables
# Testing unacceptable numeric variables
# Testing acceptable categorical variables
# Testing unacceptable categorical variables
# Testing acceptable geographic variables
# Testing unacceptable geographic variables
```

## Combining Years
# Part 2: Writing a Generic Function for Summarizing
## Generic Functions
## Numeric Summaries
## Graphical Summaries
