---
title: "Project1_SM"
author: "Sofia Mariano"
date: "30 Sept. 2026"
format: html
editor: visual
---
Prepare session to perform analysis:
```{r}
library(httr)
library(jsonlite)
library(tidyverse)
library(dplyr)
```

# Part 1: Data Processing
## Parsing Variable Values
Create list for 2021
```{r}
# Read in raw 2021 data
data_2021 <- httr::GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2021 <- jsonlite::fromJSON(rawToChar(data_2021$content))
variables_2021 <- parsed_2021$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION")]
```

Create list for 2022
```{r}
# Read in raw 2022 data
data_2022 <- httr::GET("https://api.census.gov/data/2022/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2022 <- jsonlite::fromJSON(rawToChar(data_2022$content))
variables_2022 <- parsed_2022$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION")]
```

Create list for 2023
```{r}
# Read in raw 2023 data
data_2023 <- httr::GET("https://api.census.gov/data/2023/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2023 <- jsonlite::fromJSON(rawToChar(data_2023$content))
variables_2023 <- parsed_2023$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION")]
```

Create list for 2024
```{r}
# Read in raw 2024 data
data_2024 <- httr::GET("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2024 <- jsonlite::fromJSON(rawToChar(data_2024$content))
variables_2024 <- parsed_2024$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION")]
```

Combine individual year variable lists into a single RDS file
```{r}
# Save all individual year variable lists into a single list
combined_variables <- c(variables_2021, variables_2022, variables_2023, variables_2024)

# Save combined list as a single RDS file
saveRDS(combined_variables, file = "data/rawdata_variables.RDS")
```

## Parsing API Data
Using usual coding practice, query API and return a tibble.  Data from 2021 used for this example.
```{r}
# Read-in URL
usual_2021 <- httr::GET("https://api.census.gov/data/2021/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=b195a5a7b89af4ff112dcbfe5e6a9eff2102afad")

# Parse data
parsed_usual_2021 <- jsonlite::fromJSON(rawToChar(usual_2021$content))

# Return as a tibble
tibble_2021 <- dplyr::as_tibble(parsed_usual_2021)

# Print tibble
tibble_2021
```

Create a single function that would combine all three processes above that would directly return a given URL as a tibble
```{r}
search_census <- function(URL){
  result <- httr::GET(URL)
  parsed <- jsonlite::fromJSON(rawToChar(result$content))
  return(dplyr::as_tibble(parsed))
}
```

Test function using example data from 2023.
```{r}
# Set URL variable with example data 
URL_2023 <- "https://api.census.gov/data/2023/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=20&key=b195a5a7b89af4ff112dcbfe5e6a9eff2102afad"

# Test function
search_census(URL_2023)
```

Write a function that would query an API while also parsing the values.
```{r}
query_census <- function(year, numeric, )
```

## Combining Years
# Part 2: Writing a Generic Function for Summarizing
## Generic Functions
## Numeric Summaries
## Graphical Summaries
