---
title: "Project1_SM"
author: "Sofia Mariano"
date: "30 Sept. 2026"
format: html
editor: visual
---
Prepare session to perform analysis:
```{r}
library(httr)
library(jsonlite)
library(tidyverse)
library(dplyr)
library(lubridate)
library(stringr)
```

# Part 1: Data Processing
Because the U.S. is such a large and quickly changing country, we want to look at census data to get an idea of how quickly changes have occurred throughout different parts of the U.S. within the past 2-5 years.  In order to investigate census data, we can create several functions that would allow us to look at and investigate pseudo person-level data through the Public Use Microdata Sample (PUMS) Census.  (For the case of simplicity, we're going to exclude data from Puerto Rico from our analysis)

To access the PUMS census data, we will need APIs that would help us to pull information into R and allow us to investigate the data in a programming language that we know.

Before getting to the data itself, we will need to investigate what type of census data is available from 2021-2024 and pull in the variables that are of interest to us.  In our case, we will want to look at data including people's ages (AGEP), whether or not they have had a child in the past year(FER), their monthly gas prices (GASP), a relative proportion of their monthly household income goes towards paying rent (GRPIP), what language is spoken in their household (HHL), times they arrive and leave work(JWAP and JWDP), length of their commute (JWMNP), how many people there may be in a particular row of our data (PWGTP), whether or not they are currently enrolled in school (SCH), what level of education they have (SCHL), and what region/ state of the U.S. they currently live in (REGION, DIVISION, ST, or STATE).

Knowing what information we want for each year, we're going to begin by pulling in variable lists from each individual year:

## Parsing Variable Values
Create variable list for 2021:
```{r}
# Read in raw 2021 data
data_2021 <- httr::GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2021 <- jsonlite::fromJSON(rawToChar(data_2021$content))
variables_2021 <- parsed_2021$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION")]
```

Create variable list for 2022:
```{r}
# Read in raw 2022 data
data_2022 <- httr::GET("https://api.census.gov/data/2022/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2022 <- jsonlite::fromJSON(rawToChar(data_2022$content))
variables_2022 <- parsed_2022$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION")]
```

Create variable list for 2023:
```{r}
# Read in raw 2023 data
data_2023 <- httr::GET("https://api.census.gov/data/2023/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2023 <- jsonlite::fromJSON(rawToChar(data_2023$content))
variables_2023 <- parsed_2023$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION")]
```

Create variable list for 2024:
```{r}
# Read in raw 2024 data
data_2024 <- httr::GET("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2024 <- jsonlite::fromJSON(rawToChar(data_2024$content))
variables_2024 <- parsed_2024$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION")]
```

Now that all individual year variable lists have been created, we will want to combine them into a single RDS file (R studio data file) that we can reference later on when creating our functions:
```{r}
# Save all individual year variable lists into a single list
combined_variables <- c(variables_2021, variables_2022, variables_2023, variables_2024)

# Save combined list as a single RDS file
saveRDS(combined_variables, file = "data/rawdata_variables.RDS")
```

## Parsing API Data
Looking at our previous R experience, we can start by using our usual coding practices as a template to query an API and return a tibble of the data that we want.  In this case, we're going to practice using some data from 2021.
```{r}
# Read-in URL
usual_2021 <- httr::GET("https://api.census.gov/data/2021/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=b195a5a7b89af4ff112dcbfe5e6a9eff2102afad")

# Parse data
parsed_usual_2021 <- jsonlite::fromJSON(rawToChar(usual_2021$content))

# Return as a tibble
tibble_2021 <- dplyr::as_tibble(parsed_usual_2021)

# Print tibble
tibble_2021

# The column names aren't what they should be in the first tibble.  We're going to modify the tibble output so that the first row now acts as the column names.
colnames(tibble_2021) <- tibble_2021[1,]
tibble_2021 <- tibble_2021[-1,]

# Print tibble
tibble_2021

# The output variables aren't the appropriate type for the type of data analysis we want to do further down the line.  We're going to fix variables so that their type matches the predicate type listed in the variables definition link.
tibble_2021 <- tibble_2021 |>
    dplyr::mutate(SEX <- as.factor(SEX),
                PWGTP <- as.numeric(PWGTP),
                MAR <- as.factor(MAR),
                SCHL <- as.factor(SCHL)
                ) |>
    dplyr::select(`SEX <- as.factor(SEX)`, `PWGTP <- as.numeric(PWGTP)`, `MAR <- as.factor(MAR)`, `SCHL <- as.factor(SCHL)`) |>
    dplyr::rename("SEX" = "SEX <- as.factor(SEX)", "PWGTP" = "PWGTP <- as.numeric(PWGTP)", "MAR" = "MAR <- as.factor(MAR)", "SCHL" = "SCHL <- as.factor(SCHL)")

# Print tibble to check output
tibble_2021

# The final tibble looks to have the correct column names and variable types for the data analysis we want to do further down the road.
```

Since our usual process is now working as we'd like, we will start by creating a single function to combine all of the processes that we went through in the previous coding block to allow us to directly input our URL of choice and get a tibble with the data we want as a result.
```{r}
search_census <- function(URL){
  result <- httr::GET(URL)
  parsed <- jsonlite::fromJSON(rawToChar(result$content))
  tibble_census_test <- dplyr::as_tibble(parsed)
  colnames(tibble_census_test) <- tibble_census_test[1,]
  tibble_census_test <- tibble_census_test[-1,]
  tibble_census_test <- tibble_census_test |>
    dplyr::mutate(SEX <- as.factor(SEX),
                PWGTP <- as.numeric(PWGTP),
                MAR <- as.factor(MAR),
                SCHL <- as.factor(SCHL)
                ) |>
    dplyr::select(`SEX <- as.factor(SEX)`, `PWGTP <- as.numeric(PWGTP)`, `MAR <- as.factor(MAR)`, `SCHL <- as.factor(SCHL)`) |>
    dplyr::rename("SEX" = "SEX <- as.factor(SEX)", "PWGTP" = "PWGTP <- as.numeric(PWGTP)", "MAR" = "MAR <- as.factor(MAR)", "SCHL" = "SCHL <- as.factor(SCHL)")
  return(tibble_census_test)
}
```

In order to test the function we just created above, we're going to use an example link from 2023.
```{r}
# Set URL variable with example data 
URL_2023 <- "https://api.census.gov/data/2023/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=20&key=b195a5a7b89af4ff112dcbfe5e6a9eff2102afad"

# Test function
search_census(URL_2023)
```

Since our first round of creating a function that would allow for inputting a URL and getting a tibble back worked well, we're going to now work on creating a function that would allow us to specify the variables that we want to look at and parse (clean) the values and output a tibble that contains the information we want to look at in the variable formatting that we can use to perform different analyses further down the line.

NOTE: the use of * as the geography value calls all items within the geographic value.  To subset to a specific level, you can input a number instead when specifying different inputs with this function. (you must use the formatting in the available values portion for each geography level):
Available values for region: 1-4
Available values for division: 1-9
Available values for st (years 2021-2022) or state (years 2023-2024): 01-02, 04-06, 08-13, 15-42, 44-51, 53-56
```{r}
# Call raw data variable file in order to reference in function
variable_list <- readRDS("data/rawdata_variables.RDS")

# Create function
query_census <- function(year = "2024", numeric1 = "PWGTP", numeric2 = "AGEP", categorical1 = "SEX", geography = "STATE", geography_value = "06"){
  # year variable formatting
  ifelse(year %in% c("2021","2022","2023","2024"), print(year, quote = FALSE), print("Error: Incorrect Year"))
  # numeric variable formatting
  if (numeric1 %in% c("AGEP", "GASP", "GRPIP", 
"JWMNP","PWGTP")) {
    print(numeric1, quote = FALSE)
  }
  if (numeric2 %in% c("AGEP", "GASP", "GRPIP", 
"JWMNP","PWGTP")) {
    print(numeric2, quote = FALSE)
  }
  if ((numeric1 != "PWGTP") & (numeric2 != "PWGTP")){
    "Error: Missing Key Numeric Variable PWGTP"
  }
  #categorical variable formatting
  ifelse(categorical1 %in% c("FER", "HHL", "SCH", "SCHL", "SEX"), print(categorical1, quote = FALSE), print("Error: Incorrect categorical variable specified"))
  # geographic variable formatting
  ifelse(geography %in% c("REGION", "DIVISION", "ST", "STATE"), print(geography, quote = FALSE), print("Error: Incorrect Geography Level"))
    if (geography_value == "*"){ 
    print(geography_value, quote = FALSE)
  } else if (((geography == "REGION") & (geography_value %in% c("1", "2", "3", "4"))) | ((geography == "DIVISION") & (geography_value %in% c("1", "2", "3", "4", "5", "6", "7", "8", "9"))) | ((geography == "ST") & (geography_value %in% c("01", "02", "04", "05", "06", "08", "09", "10", "11", "12", "13", "15", "16", "17", "18", "19", "20","21", "22", "23", "24","25", "26", "27", "28", "29", "30", "31", "32", "33", "34", "35", "36", "37", "38", "39", "40", "41", "42", "44", "45", "46", "47", "48","49", "50", "51", "53", "54", "55", "56", "72"))) | ((geography == "STATE") & (geography_value %in% c("01", "02", "04", "05", "06", "08","09", "10", "11", "12", "13", "15", "16", "17", "18", "19", "20", "21", "22", "23", "24", "25", "26", "27", "28", "29", "30", "31", "32", "33", "34", "35", "36", "37", "38", "39", "40", "41", "42", "44", "45", "46", "47", "48", "49", "50", "51", "53", "54", "55", "56", "72")))){
    print(geography_value, quote = FALSE)
    } else {print("Error: Incorrect Geography Subset Specified")}

# coding for URL begins after formatting completed
  result <- httr::GET(paste0("https://api.census.gov/data/",
                             year,
                             "/acs/acs1/pums?get=",
                             numeric1,
                             ",",
                             numeric2,
                             ",",
                             categorical1,
                             "&for=",
                             geography,
                             ":",
                             geography_value,
                             "&key=b195a5a7b89af4ff112dcbfe5e6a9eff2102afad"
  ))
  parsed <- jsonlite::fromJSON(rawToChar(result$content))
  tibble_census <- dplyr::as_tibble(parsed)
  colnames(tibble_census) <- tibble_census[1,]
  tibble_census <- tibble_census[-1,]
  tibble_census <- tibble_census |>
    dplyr::mutate(PWGTP <- as.numeric(PWGTP),
                  if(numeric1 == "AGEP") {AGEP <- as.numeric(AGEP)},
                  if (numeric2 == "AGEP") {AGEP <- as.numeric(AGEP)},
                  if((numeric1 == "GASP") | (numeric2 == "GASP")) {GASP <- as.numeric(GASP)},
                  if((numeric1 == "GRPIP") | (numeric2 == "GRPIP")) {GRPIP <- as.numeric(GRPIP)},
                  if((numeric1 == "JWMNP") | (numeric2 == "JWMNP")) {JWMNP <- as.numeric(JWMNP)},
                  if((numeric1 == "JWAP") | (numeric2 == "JWAP")) {JWAP <- for (item in variable_list$JWAP$values$item){
    if (is.na(item)){
      "NA"
    } else if (!is.na(item)){
            #select out N/A values for 0
      if (item != "to"){
        print(item)
      } else {calculate_midpoint <- function(item)
      # Split string into start and end time
      time_parts <- stringr::str_split(item, "to")[]
    # Parse individual times
    start_time <- readr::parse_time(time_parts)
    end_time <- readr::parse_time(time_parts)
    # Convert seconds since midnight
    start_seconds <- as.numeric(start_time) %% (24*60*60)
    end_seconds <- as.numeric(end_time) %% (24*60*60)
    # Calculate midpoint in seconds
    midpoint_seconds <- (start_seconds + end_seconds) / 2
    # Convert midpoint to time object
    midpoint_time <- as.POSIXct(midpoint_seconds, tz = "UTC")
    # Format to display time only
    midpoint_format <- format(midpoint_time, format = "%I:%M %p")
    return(midpoint_format)
      }
    }
  }
                  },
                  if((numeric1 == "JWDP") | (numeric2 == "JWDP")) {JWDP <- for (item in variable_list$JWDP$values$item){
    if (is.na(item)){
      "NA"
    } else if (!is.na(item)){
      #select out N/A values for 0
      if (item != "to"){
        print(item)
      } else {calculate_midpoint <- function(item)
      # Split string into start and end time
      time_parts <- stringr::str_split(item, "to")[]
    # Parse individual times
    start_time <- readr::parse_time(time_parts)
    end_time <- readr::parse_time(time_parts)
    # Convert seconds since midnight
    start_seconds <- as.numeric(start_time) %% (24*60*60)
    end_seconds <- as.numeric(end_time) %% (24*60*60)
    # Calculate midpoint in seconds
    midpoint_seconds <- (start_seconds + end_seconds) / 2
    # Convert midpoint to time object
    midpoint_time <- as.POSIXct(midpoint_seconds, tz = "UTC")
    # Format to display time only
    midpoint_format <- format(midpoint_time, format = "%I:%M %p")
    return(midpoint_format)
      }
    }
  }
                  },
                  if(categorical1 == "FER") {FER <- factor(FER, 0:2, print(variable_list$FER$values$item))},
                  if(categorical1 == "SEX") {SEX <- factor(SEX, 1:2, print(variable_list$SEX$values$item))},
                  if(categorical1 == "SCH") {SCH <- factor(SCH, 0:3, print(variable_list$SCH$values$item))},
                  if(categorical1 == "SCHL") {SCHL <- factor(SCHL, 0:24, print(variable_list$SCH$values$item))}
    )
  class(tibble_census) <- c("census", class(tibble_census))
  return(tibble_census)
}
```

Testing the function in various conditions. 
NOTE: add quotations when putting in different values for inputs
```{r}
# Testing defaults
query_census()

# Testing acceptable years
query_census(year = "2023")

# Testing unacceptable categorical variable
query_census(year = "2022", categorical1 = "CIT")
```

Since all of the outputs are working as we'd like them to, we can move forward with creating more complex functions.

## Combining Years
If we want to search for multiple years at once, we're going to have to slightly modify the function above:
```{r}
# First, we should save the results of each function for all of our individual years.  Because we'll be combining multiple years, we're going to add a "years" column that would allow us to trace back which year we're pulling data from (called year).  For simplicity, we're going to switch our geography variable to region and set it's value at 4 (West)
census_2021 <- query_census(year = "2021", geography = "REGION", geography_value = "4")
census_2021$year <- "2021"

census_2022 <- query_census(year = "2022", geography = "REGION", geography_value = "4")
census_2022$year <- "2022"

census_2023 <- query_census(year = "2023", geography = "REGION", geography_value = "4")
census_2023$year <- "2023"

census_2024 <- query_census(year = "2024", geography = "REGION", geography_value = "4")
census_2024$year <- "2024"

# Now we need to combine this data into a single final tibble that we can call census_all
census_all <- dplyr::bind_rows(census_2021, census_2022, census_2023, census_2024)

# Look at the results:
census_all
```

# Part 2: Writing a Generic Function for Summarizing
## Generic Functions
## Numeric Summaries
## Graphical Summaries
