---
title: "Project1_SM"
author: "Sofia Mariano"
date: "30 Sept. 2026"
format: html
editor: visual
---
Prepare session to perform analysis:
```{r}
library(httr)
library(jsonlite)
library(tidyverse)
library(dplyr)
library(lubridate)
library(stringr)
```

# Part 1: Data Processing
## Parsing Variable Values
Create list for 2021
```{r}
# Read in raw 2021 data
data_2021 <- httr::GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2021 <- jsonlite::fromJSON(rawToChar(data_2021$content))
variables_2021 <- parsed_2021$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION")]
```

Create list for 2022
```{r}
# Read in raw 2022 data
data_2022 <- httr::GET("https://api.census.gov/data/2022/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2022 <- jsonlite::fromJSON(rawToChar(data_2022$content))
variables_2022 <- parsed_2022$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION")]
```

Create list for 2023
```{r}
# Read in raw 2023 data
data_2023 <- httr::GET("https://api.census.gov/data/2023/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2023 <- jsonlite::fromJSON(rawToChar(data_2023$content))
variables_2023 <- parsed_2023$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION")]
```

Create list for 2024
```{r}
# Read in raw 2024 data
data_2024 <- httr::GET("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2024 <- jsonlite::fromJSON(rawToChar(data_2024$content))
variables_2024 <- parsed_2024$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION")]
```

Combine individual year variable lists into a single RDS file
```{r}
# Save all individual year variable lists into a single list
combined_variables <- c(variables_2021, variables_2022, variables_2023, variables_2024)

# Save combined list as a single RDS file
saveRDS(combined_variables, file = "data/rawdata_variables.RDS")
```

## Parsing API Data
Using usual coding practice, query API and return a tibble.  Data from 2021 used for this example.
```{r}
# Read-in URL
usual_2021 <- httr::GET("https://api.census.gov/data/2021/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=b195a5a7b89af4ff112dcbfe5e6a9eff2102afad")

# Parse data
parsed_usual_2021 <- jsonlite::fromJSON(rawToChar(usual_2021$content))

# Return as a tibble
tibble_2021 <- dplyr::as_tibble(parsed_usual_2021)

# Print tibble
tibble_2021

# Column names aren't what they should be in the first tibble.  Fix tibble output so the first row now acts as the column names.
colnames(tibble_2021) <- tibble_2021[1,]
tibble_2021 <- tibble_2021[-1,]

# Print tibble
tibble_2021

# The output variables aren't the appropriate type.  Fix variables so that their type matches the predicate type listed in the variables definition link.
tibble_2021 <- tibble_2021 |>
    dplyr::mutate(SEX <- as.factor(SEX),
                PWGTP <- as.numeric(PWGTP),
                MAR <- as.factor(MAR),
                SCHL <- as.factor(SCHL)
                ) |>
    dplyr::select(`SEX <- as.factor(SEX)`, `PWGTP <- as.numeric(PWGTP)`, `MAR <- as.factor(MAR)`, `SCHL <- as.factor(SCHL)`) |>
    dplyr::rename("SEX" = "SEX <- as.factor(SEX)", "PWGTP" = "PWGTP <- as.numeric(PWGTP)", "MAR" = "MAR <- as.factor(MAR)", "SCHL" = "SCHL <- as.factor(SCHL)")

# Print tibble to check output
tibble_2021
```

Create a single function that would combine all three processes above that would directly return a given URL as a tibble
```{r}
search_census <- function(URL){
  result <- httr::GET(URL)
  parsed <- jsonlite::fromJSON(rawToChar(result$content))
  tibble_census_test <- dplyr::as_tibble(parsed)
  colnames(tibble_census_test) <- tibble_census_test[1,]
  tibble_census_test <- tibble_census_test[-1,]
  tibble_census_test <- tibble_census_test |>
    dplyr::mutate(SEX <- as.factor(SEX),
                PWGTP <- as.numeric(PWGTP),
                MAR <- as.factor(MAR),
                SCHL <- as.factor(SCHL)
                ) |>
    dplyr::select(`SEX <- as.factor(SEX)`, `PWGTP <- as.numeric(PWGTP)`, `MAR <- as.factor(MAR)`, `SCHL <- as.factor(SCHL)`) |>
    dplyr::rename("SEX" = "SEX <- as.factor(SEX)", "PWGTP" = "PWGTP <- as.numeric(PWGTP)", "MAR" = "MAR <- as.factor(MAR)", "SCHL" = "SCHL <- as.factor(SCHL)")
  return(tibble_census_test)
}
```

Test function using example data from 2023.
```{r}
# Set URL variable with example data 
URL_2023 <- "https://api.census.gov/data/2023/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=20&key=b195a5a7b89af4ff112dcbfe5e6a9eff2102afad"

# Test function
search_census(URL_2023)
```

Write a function that would query an API while also parsing the values.
NOTE: the use of * as the geography value calls all items within the geographic value.  To subset to a specific level, you can input a number instead when specifying different inputs with this function.
```{r}
# Call raw data variable file in order to reference in function
variable_list <- readRDS("data/rawdata_variables.RDS")

# Create function
query_census <- function(year = "2024", numeric1 = "PWGTP", numeric2 = "AGEP", categorical1 = "SEX", geography = "ST", geography_value = "*",key = "b195a5a7b89af4ff112dcbfe5e6a9eff2102afad"){
  # year variable formatting
  ifelse(year %in% c("2021","2022","2023","2024"), print(year), print("Error: Incorrect Year"))
  # numeric variable formatting
  if (numeric1 %in% c("AGEP", "GASP", "GRPIP", 
"JWMNP","PWGTP")) {
    print(numeric1)
  }
  if (numeric2 %in% c("AGEP", "GASP", "GRPIP", 
"JWMNP","PWGTP")) {
    print(numeric2)
  }
  if ((numeric1 != "PWGTP") & (numeric2 != "PWGTP")){
    "Error: Missing Key Numeric Variable PWGTP"
  }
  #categorical variable formatting
  ifelse(categorical1 %in% c("FER", "HHL", "SCH", "SCHL", "SEX"), print(categorical1), print("Error: Incorrect categorical variable specified"))
  # geographic variable formatting
  ifelse(geography %in% c("REGION", "DIVISION", "ST", "STATE"), print(geography), print("Error: Incorrect Geography Level"))
  if ((geography == "REGION") & (geography_value %in% variable_list$REGION$values)) {
    print(geography_value)
  } else if ((geography == "DIVISION") & (geography_value %in% variable_list$DIVISION$values)) {
    print(geography_value)
  } else if ((geography == "ST") & (geography_value %in% variable_list$ST$values)) {
    print(geography_value)
  } else if ((geography == "STATE") & (geography_value %in% variable_list$STATE$values)){
    print(geography_value)
  } else {
    "Error: Incorrect Geography Subset Specified"
  }

# coding for URL begins after formatting completed
  result <- httr::GET(paste0("https://api.census.gov/data/",
                             year,
                             "/acs/acs1/pums?get=",
                             numeric1,
                             ",",
                             numeric2,
                             ",",
                             categorical1,
                             "&for",
                             geography,
                             ":",
                             geography_value,
                             "&key=",
                             key
  ))
  parsed <- jsonlite::fromJSON(rawToChar(result$content))
  tibble_census <- dplyr::as_tibble(parsed)
  colnames(tibble_census) <- tibble_census[1,]
  tibble_census <- tibble_census[-1,]
  tibble_census <- tibble_census |>
    dplyr::mutate(PWGTP <- as.numeric(PWGTP),
                  AGEP <- as.numeric(AGEP),
                  GASP <- as.numeric(GASP),
                  GRPIP <- as.numeric(GRPIP),
                  JWMNP <- as.numeric(JWMNP),
                  JWAP <- for (item in variable_list$JWAP$values$item){
    if (is.na(item)){
      "NA"
    } else if (!is.na(item)){
            #select out N/A values for 0
      if (item != "to"){
        print(item)
      } else {calculate_midpoint <- function(item)
      # Split string into start and end time
      time_parts <- stringr::str_split(item, "to")[]
    # Parse individual times
    start_time <- readr::parse_time(time_parts)
    end_time <- readr::parse_time(time_parts)
    # Convert seconds since midnight
    start_seconds <- as.numeric(start_time) %% (24*60*60)
    end_seconds <- as.numeric(end_time) %% (24*60*60)
    # Calculate midpoint in seconds
    midpoint_seconds <- (start_seconds + end_seconds) / 2
    # Convert midpoint to time object
    midpoint_time <- as.POSIXct(midpoint_seconds, tz = "UTC")
    # Format to display time only
    midpoint_format <- format(midpoint_time, format = "%I:%M %p")
    return(midpoint_format)
      }
    }
  },
                  JWDP <-   for (item in variable_list$JWDP$values$item){
    if (is.na(item)){
      "NA"
    } else if (!is.na(item)){
      #select out N/A values for 0
      if (item != "to"){
        print(item)
      } else {calculate_midpoint <- function(item)
      # Split string into start and end time
      time_parts <- stringr::str_split(item, "to")[]
    # Parse individual times
    start_time <- readr::parse_time(time_parts)
    end_time <- readr::parse_time(time_parts)
    # Convert seconds since midnight
    start_seconds <- as.numeric(start_time) %% (24*60*60)
    end_seconds <- as.numeric(end_time) %% (24*60*60)
    # Calculate midpoint in seconds
    midpoint_seconds <- (start_seconds + end_seconds) / 2
    # Convert midpoint to time object
    midpoint_time <- as.POSIXct(midpoint_seconds, tz = "UTC")
    # Format to display time only
    midpoint_format <- format(midpoint_time, format = "%I:%M %p")
    return(midpoint_format)
      }
    }
  },
                  FER <- factor(FER, 0:2, print(variable_list$FER$values$item)
                     ),
                  SEX <- factor(SEX, 1:2, print(variable_list$SEX$values$item)
                     ),
                  SCH <- factor(SCH,0:3, print(variable_list$SCH$values$item)
                     ),
                  SCHL <- factor(SCHL,0:24, print(variable_list$SCH$values$item)
                      )
                )
  class(parsed) <- c("census", class(parsed))
  return(tibble_census)
}
```

Testing the function in various conditions
```{r}
# Testing defaults
query_census()

# Testing acceptable years
query_census("2023")

# Testing unacceptable geographic variables
```

## Combining Years
# Part 2: Writing a Generic Function for Summarizing
## Generic Functions
## Numeric Summaries
## Graphical Summaries
