---
title: "Project1_SM"
author: "Sofia Mariano"
date: "30 Sept. 2026"
format: html
editor: visual
---
Prepare session to perform analysis:
```{r}
library(httr)
library(jsonlite)
library(tidyverse)
library(dplyr)
```

# Part 1: Data Processing
## Parsing Variable Values
Create list for 2021
```{r}
# Read in raw 2021 data
data_2021 <- httr::GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2021 <- jsonlite::fromJSON(rawToChar(data_2021$content))
variables_2021 <- parsed_2021$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION")]

# Save the list as an RDS file in the data folder
saveRDS(variables_2021, file = "data/rawdata_2021.RDS")
```

Create list for 2022
```{r}
# Read in raw 2022 data
data_2022 <- httr::GET("https://api.census.gov/data/2022/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2022 <- jsonlite::fromJSON(rawToChar(data_2022$content))
variables_2022 <- parsed_2022$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION")]

# Save list as an RDS file in the data folder
saveRDS(variables_2022, file = "data/rawdata_2022.RDS")
```

Create list for 2023
```{r}
# Read in raw 2023 data
data_2023 <- httr::GET("https://api.census.gov/data/2023/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2023 <- jsonlite::fromJSON(rawToChar(data_2023$content))
variables_2023 <- parsed_2023$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION")]

# Save the list as an RDS file in the data folder
saveRDS(variables_2023, file = "data/rawdata_2023.RDS")
```

Create list for 2024
```{r}
# Read in raw 2024 data
data_2024 <- httr::GET("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")

# Parse and get a list of the desired variables
parsed_2024 <- jsonlite::fromJSON(rawToChar(data_2024$content))
variables_2024 <- parsed_2024$variables[c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER", "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION")]

# Save the list as an RDS file in the data folder
saveRDS(variables_2024, file = "data/rawdata_2024.RDS")
```

## Parsing API Data
Using usual coding practice, query API and return a tibble.  Data from 2021 used for this example.
```{r}
# Read-in URL
usual_2021 <- httr::GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

# Parse data
parsed_usual_2021 <- jsonlite::fromJSON(rawToChar(usual_2021$content))

# Return as a tibble
tibble_2021 <- dplyr::as_tibble(parsed_usual_2021$data)

# Print results to observe if desired output produced
tibble_2021

```

## Combining Years
# Part 2: Writing a Generic Function for Summarizing
## Generic Functions
## Numeric Summaries
## Graphical Summaries
